# Exercises for week 5

## The Human dataset

I will start off by exploring the structure and dimensions of my 'human' dataset.
```{r human dataset}
human <- read.table("/Users/Noora/Documents/IODS-project/data/create_human_.txt", header = TRUE)
dim(human)
str(human)
```
From the outputs I see that the dataset has 155 observation and 8 variables. However, I know that the row names denote the variable 'Country', thus the dataset has in reality 9 variables, not 8. For example the last row of my dataset is:
```{r last row human}
library(knitr)
knitr::kable(tail(human, n = 1))
```

Here I have 8 named columns, but the first column shows the name of the country in question (in this case Niger), so in effect there are 9 variables.

The original 'human' dataset includes variables that relate to the Human Development Index (HDI), and to other indexes such as Gender Development Index (GDI), and combines indicators from most countries. The dataset originally consisted of more than the 9 variables shown in my dataset, but it has been  wrangled and reduced to 9 variables, and to only obsrvations that relate to countries, and it includes only those observations without missing values. The variable 'Country' is the name of the country in question. The other 8 variables in my dataset relate to two topics: Health and knowledge, and Empowerment. The variables related to Empowerment in my dataset are:

+ **High_educ_f_m**: Proportion of females with at least secondary education / Proportion of males with at least secondary education
+ **F_m_working**: Proportion of females in the labour force / Proportion of males in the labour force
+ **Parl_seats**: Percetange of female representatives in parliament

The variables related to Health and knowledge in my dataset are:

+ **Educ_exp**: Expected years of schooling
+ **Life_exp**: Life expectancy at birth
+ **GNI_**: Gross National Income per capita
+ **Mat_deaths**: Maternal mortality ratio
+ **Young_births**: Adolescent birth rate

The source for the definitions of the variables is [this website](https://raw.githubusercontent.com/TuomoNieminen/Helsinki-Open-Data-Science/master/datasets/human_meta.txt).

## Overview of the data

Here is a graphical overview of the data:
``` {r graphical human}
library(ggplot2)
library(GGally)
GraphOV <- ggpairs(human, mapping = aes(), title = "Graphical Overview", lower = list(combo = wrap("facethist", bins = 20)))
GraphOV
```

Here are also summaries of the variables in my dataset (the knitr part of the code just formats the result into a nice table):
``` {r human summaries}
knitr::kable(summary(human))
```

From the summary table I see immediately that many of the variables have fairly large ranges. For example some countries have no women at all in parliament (as the min value of Parl_seats is 0), whereas some have more women in parliament than men (as the max value is above 50%). This already suggests to me that some countries in the dataset are quite different from each other. Similarly there seems to be large differences between some countries with regard to all the other variables. For example in some countries the expected years of schooling are under 6 and in others above 20. At the same time, expected years of schooling seems to be among those variables where many countries also are fairly similar, with 50% of countries expecting inhabitants to have between 11 and 15 years of schooling. This variable is also the one that seems to be the closest to being normally distributd from all the variables when looking at the graphical overview.

The graphical overview indicates that some of the variables may be related to one another. For example it seems that as adolescent birth rate increase, so does maternal mortality ratio. Not surprisingly, there also seems to be a relationship between maternal mortality ratio and life expectancy, with higher maternal mortality ratios corresponding to lower lif expectancies. In fact, a great number of the variables seem to be correlated. Another example of such variables are education and life expectancies: the higher the education expectancy, the higher also the life expectancy and vice versa. Among those variable pairs that don't seem to be strongly correlated are e.g. expected years of schooling and percetange of female representatives in parliament.

## Principal component analysis (PCA)

Next I will perform principal component analysis (PCA) without standardising the dataset. First I will look at the variability captured by the principal components. I will also draw a biplot displaying the observations with PC1 in coordinate x-axis and PC2 in y-axis.

``` {r human not standardised}
pca_human <- prcomp(human)
summary(pca_human)
biplot(pca_human, choices = 1:2, cex = c(1, 1), col = c("grey40", "deeppink2"), sub = "Plot 1: PC1 & PC2 with non standardised dataset")
```

However, I should first standardise the variables to be able to conduct analysis, so I repeat the PCA, and look at the variability and biplot of the tandardised dataset next.

``` {r human standardised PCA}
human_std <- scale(human)
pca_human_std <- prcomp(human_std)
summary(pca_human_std)
biplot(pca_human_std, choices = 1:2, cex = c(1, 1), col = c(1, 14), sub = "Plot 2: PC1 & PC2 with standardised dataset")
```

The results with and without standardising are very different. When the dataset is not standardised, PC1 explains nearly all of the variance (99.99%) in the original variables, thus PC2 contribute very little and the other PCs not at all toward explaining the variance. From plot 1 I can also see that there is only one very visible arrow ('GNI_'), which is parallel to x-axis, meaning that GNI has high correlation with PC1. The reason why GNI is so dominant is because in the non standardised dataset GNI has much higher values than any of the other variables, which is why standardisation of the dataset is needed.GNI also has clearly the largest standard deviation in this model.

With standardising, PC1 still explains most of the variance in the original variables, with its proportion of variance being 54%. All 8 PCs are needed to get explain all of the variance in the original variables. In plot 2 I see all the variables mapped out clearly, with none of the variables having noticably greater standard deviation than the others.The arrows point to different directions, with some variables being correlated with PC1 whilst others with PC2 (see section below). In the second plot countries are also more evenly distributed in the scatter plot, whereas in picture 1 they were mainly scattered to the top right cornr of the plot.

## Interpreting PC1 and PC2

My personal interpretation of the first two principal component dimensions (PC1 and PC2) based on the biplot of the standardised dataset is that PC1 relates (mainly) to the topic 'Health and knowledge' and PC2 relates to the topic 'Empowerment'. The two 'Empowerment' related variables 'F_m_working' and 'Parl_seats' are fairly well parallel with the y-axis (have high correlation with PC2), whereas all the other variables are somewhat parallel with the x-axis (have high correlation with PC1). However, some of the variables seem to be positively related to PC1 and some negatively. 'Mat_deaths' and 'Young_births' arrows point to the right, whereas the education-related variables ('Educ_exp', 'High_educ_f_m'), 'GNI_' and 'Life_exp' point to the opposite direction. It is not surprising that for example life expectancy and maternal mortality ratio arrows point to the opposite directions (high negative correlation), as high maternal mortality clearly must negatively relate to life expectancy. The 'Health and knowledge' and 'Empowerment' related variables (with the exception of 'High_educ_f_m') are (almost) orthogonal, meaning that they are uncorrelated with each other.

## The tea dataset

Next I move to another dataset - the tea dataset from the FactoMineR package.

``` {r loading tea}
library(FactoMineR)
data("tea")
str(tea)
dim(tea)
```

The dataset seems to have 36 variables and 300 observations. The variables are e.g. age, sex, lunch and dinner, and seem to relate to tea consumption. I want to only keep some interesting variables in my analysis, so I choose to focus on variables:

+ frequency: '1/day', '1 to 2/week', '+2/day' and '3 to 6/week'
+ Tea: 'Earl Grey', 'green' and 'black'
+ sugar: 'sugar' and 'no sugar'
+ tea.time: 'not tea time' and 'tea time'
+ sex: 'female' (F) and 'male' (M)

I will next visualise these variables from the dataset.

``` {r plot tea}
library(tidyr)
library(dplyr)
keep_columns <- c("frequency", "Tea", "sugar", "tea.time", "sex")
te1 <- dplyr::select(tea, one_of(keep_columns))
gather(te1) %>% ggplot(aes(value)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + facet_wrap("key", scales = "free")
```

## Multiple correspondence analysis and the tea data

Next I will perform MCA on the variables of the tea dataset that I kept for analysis.

``` {r mca tea}
mca_tea <- MCA(te1, graph = FALSE)
summary(mca_tea)
```

From the Eigenvalues I see that the first dimensions retains 20% of the total variances, whilst the second one retains 15%. The differences of how much variance the 8 dimensions retain are fairly small (e.g. dimensions 4 retains 12.6% and dimensions 5 11.8%). The categories section shows the v-test results for the dimensions. Most of v-test values are above 1.96 or below -1.96, suggesting that the coorsidinates are significantly different from 0, the only exception being the frequency-variable's values. None of the variables seem to be strongly linked with a specific dimension (squared correlation between the variables and dimensions is not very close to 1), with perhaps the type of tea (especially Earl Grey) and dimension 2 being the most strongly linked.

``` {r visualisation mca tea}
plot(mca_tea, invisible = c("ind"), habillage = "quali", sub = "Plot 3: MCA of tea dataset")
```

The MCA biplot shows how similar the variables are with each other. It seems that 'tea time' and 'female' are very similar, and 'not tea time' and 'male'. The type of tea most similar with 'sugar' is 'Earl Grey' whereas 'black' is most similar with 'no sugar'. It also seems that 'green' is quite different from the other variables (maybe most similar to it is 'black'). Frequent consumption '+2 / day' is more similar to 'female' and 'no sugar' than to 'male' and 'sugar' whereas less frequent uses (e.g. '1 to 2 / week') are more similar with 'male' and 'sugar'. Both 'male' and 'female' are approximately as far from 'no sugar' and 'sugar'.